### 2020-09-06 discussion.tex
final draft of popl

\section{Discussion and Conclusion}\label{sec:discussion}\label{sec:conclusion}
\begin{quote}
  %
  \textit{
  %
  The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.''
  %
  }

@@

  Another direction for future work has to do with pushing edits from computed results
  back into livelits. For example, a slider expands to a number, which may 
  then flow through a computation. Bidirectional evaluation techniques may allow
  the user to edit a number in the result of a computation and see the necessary
  change to a slider in the program.

>>>>

\section{Discussion and Conclusion}\label{sec:discussion}\label{sec:conclusion}
\begin{quote}
  %
  \textit{
  %
  The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.
  %
  }

@@

  Another direction for future work has to do with pushing edits from computed results
  back into livelits. For example, a slider expands to a number, which may 
  then flow through a computation. Bidirectional evaluation techniques may allow
  the user to edit a number in the result of a computation and see the necessary
  change to a slider in the program \cite{sns-pldi,sns-uist}.

### 2020-09-06 implementation.tex
final draft of popl

\subsection{Hazel}
Hazelnut Live is the formal foundation for the Hazel programming environment, so 
it was natural to choose the the Hazel programming environment for our primary implementation\todo{cite?}{}. 
 Hazel is a browser-based live functional programming environment 
organized from the ground up around typed holes. Uniquely, every editor state in Hazel is 
semantically meaningful: it has a type, it can be evaluated, and it can be transformed 
in a type-aware manner. This implies that livelits remain fully functional at all times, 
even when the program is incomplete or erroneous.

Hazel is implemented in OCaml and compiled to JavaScript using the \li{js_of_ocaml} compiler\todo{cite}.
Hazel is written in a pure functional style, so our implementation is able to closely 
follow the formalism. The Hazel language has been evolving in parallel with our work, 
and has only recently become expressive enough for it to be feasible to implement livelits within Hazel itself.
The examples in this paper are instead implemented in OCaml as modules satisfying the signature 
implied by Fig.~\ref{fig:color-impl}. A functor then adapts these OCaml-based livelit implementations 
to livelit definitions suitable for inclusion in the livelit context used in the semantics.

@@

\subsection{Text Editor Integration}
Before delving into these issues, let us note that livelits do not require the use of a structure editor. 
We have also developed a 
proof-of-concept textual program editor, based on Sketch-n-Sketch \cite{sns-pldi,sns-uist}, with support for livelits.
This proof-of-concept is not at feature parity with 
our main implementation, but it demonstrates that a text editor with support for 
parsing out livelit applications and replacing them with interactive graphical inclusions is sufficient.

\subsection{Layout}\label{sec:layout}
Whether implemented in a structure editor or a text editor, livelits present 
interesting layout challenges: it can be difficult to sensibly 
interleave symbolic code with livelit views.

Hazel uses an optimizing pretty printer to determine layout, an approach 
based on the work of Bernardy et al \todo{cite bernardy}. This system relies 
fundamentally on character counts. Consequently, our implementation asks each 
livelit to specify dimensions in terms of character counts rather than pixels.
Livelits can be laid out in one of two ways: as inline livelits, like \li{\$slider},
which must be exactly one character high and appear inline with the code,
or as \emph{multi-line livelits}, which occupy up to the full width of the editor 
and a specified number of code lines (not shown in Fig.~\ref{fig:color-impl}). 
The pretty printer lays out multi-line livelits
on their own line, as seen in the figures in this paper.

@@

Many livelit-related computations can be intelligently memoized, e.g. view computations and unchanged splice evaluations,
especially with the edit awareness that Hazel provides. 
We leave further optimizations along these lines to future work.
Prior work on incrementally updating UIs as code changes could be relevant to this effort \cite{burckhardt2013s}.

>>>>

\subsection{Hazel}
Hazelnut Live is the formal foundation for the Hazel programming environment, so 
it was natural to choose the the Hazel programming environment for our primary implementation. 
 Hazel is a browser-based live functional programming environment 
organized from the ground up around typed holes. Uniquely, every editor state in Hazel is 
semantically meaningful: it has a type, it can be evaluated, and it can be transformed 
in a type-aware manner. This implies that livelits remain fully functional at all times, 
even when the program is incomplete or erroneous.

Hazel is implemented in OCaml and compiled to JavaScript using the \li{js_of_ocaml} compiler \cite{DBLP:conf/aplas/RadanneVB16}.
The core of Hazel is written in a pure functional style, so our implementation is able to closely 
follow the formalism. The Hazel language has been evolving in parallel with our work, 
and has only recently become expressive enough for it to be feasible to implement livelits within Hazel itself.
The examples in this paper were instead implemented in OCaml as modules satisfying the signature 
implied by Fig.~\ref{fig:color-impl}. A functor then adapts these OCaml-based livelit implementations 
to livelit definitions suitable for inclusion in the livelit context used in the semantics.

@@

\subsection{Text Editor Integration}
Before delving into layout issues, let us note that livelits do not require the use of a structure editor. 
We have also developed a 
proof-of-concept textual program editor, based on Sketch-n-Sketch \cite{sns-pldi,sns-uist}, with support for livelits.
This proof-of-concept is not at feature parity with 
our main implementation, but it demonstrates that a syntax-recognizing text editor \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84} 
is sufficient to support livelits, albeit with some gaps in livelit UI availability when 
there are critical errors.

\subsection{Layout}\label{sec:layout}
Whether implemented in a structure editor or a text editor, livelits present 
interesting layout challenges--it can be difficult to sensibly 
interleave symbolic code with livelit views in a clean, readable manner.

Hazel uses an optimizing pretty printer to determine layout, an approach 
based on the work of \citet{DBLP:journals/pacmpl/Bernardy17}. This system relies 
fundamentally on character counts. Consequently, our implementation asks each 
livelit to specify dimensions in terms of character counts rather than pixels.
Livelits can be laid out in one of two ways: as inline livelits, like \li{\$slider},
which must be exactly one character high and appear inline with the code,
or as \emph{multi-line livelits}, which occupy up to the full width of the editor 
and a specified number of code lines (we omitted these specifications in Fig.~\ref{fig:color-impl}). 
The pretty printer lays out multi-line livelits
on their own line, as seen in Fig.~\ref{fig:grading}.

@@

Many livelit-related computations can be intelligently memoized, e.g. view computations and unchanged splice evaluations,
especially with the edit awareness that Hazel provides. 
We leave further optimizations along these lines to future work.
Prior work on incrementally updating UIs after code changes could be relevant to this effort \cite{burckhardt2013s}.
